Time slice analyses

### 2017-11-27 analyses/02-time-slice-analyses.Rmd
Remove examining output section and save otuputs instead

## Load functions using source
# Doesn't always work... Rmd problem?
source("../functions/simulations_fun.R")
```

@@

To run `dispRity` analyses we need the **morphospace**, the **phylogeny** (with a root age and node labels) and the **first and last occurrence dates** (FADLAD). The tree and FADLAD are in the `data/` folder. To construct the morphospace we need to use the `distance_matrix` files we created in `01-extract-data-for analyses.Rmd` and stored in `data/processed`.

@@

## Loading the discrete morphological matrix
## This is just to match the names with the tree
matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))

@@

## Loading the tree
tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
## Adding node labels and a root age (max tree age date)
tree <- makeNodeLabel(tree, method = "number", prefix = "n")
tree$root.time <- max(tree.age(tree)[, 1])

## Cleaning the tree to remove things that aren't in the matrix 
cleaned_data <- clean.data(matrix$matrix, tree)
matrix$matrix <- cleaned_data$matrix
tree <- cleaned_data$tree

## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

## Ordinating the matrix
morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points

```

@@

These analyses use a series of functions in `functions\`. These essentially wrap the existing `dispRity` functions so that running lots of models and extracting lots of data is not such a big issue.

> TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
We now only calculate the disparity for the non-rarefied subsets and the subsets rarefied to contain only 3 taxa (the lowest level of rarefaction).
We can use the full rarefaction if we want to plot rarefaction curves later on.

```{r, eval = TRUE, warning = FALSE}
## Run the function for Beck data
slug <- "BeckLee"

## Ages (rarefied to 3 taxa only)
out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
                                bootstraps = 100, metric = c(sum, variances),
                                metric.name = "sum_var", rarefaction = FALSE)
#readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))

## Epochs (rarefied to 3 taxa only)
out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
                        rarefaction = FALSE)
#readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
```

Note that you can use the previous output format (the data tables) as before by accessing them through `out$results$stratigraphy/duration/number`.
Writing them as a csv will require some thinking since they don't have the same number of columns (though it's still doable I guess).
Also, if you want to plot things through `ggplot`, you can use the values in these data tables!

Examining the output 
--------------------

The results as a table

```{r, eval = TRUE, echo = FALSE}
knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic time: unequal time bins and non-equidistant time slices")
knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
```

This is not really digest and we should find a better way to present that.
Anyway, these boring tables are always good for the supplementaries.

## Plotting the results `dispRity` style

Here is an example of the results using `plot.dispRity`.
The upper panel is with the CIs and the lower one is the same without the CIs.

### The results for the Age type

```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
op <- par(bty = "n", mfrow = c(3,2))
## Principal colors
colors <- palette()[1:5][-1] #TG: that can be made fancier
colors[5] <- "black"
## Color combination (for the CI)
colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
## y limits
ylim <- range(unlist(lapply(out1$results, function(X) return(c(min(X[,5], na.rm = TRUE), max(X[,5],
  na.rm = TRUE))))))
## Zooming
ylim_zoom <- c(20, 35)

## Plotting the different methods (with/without CIs)
plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration", ylim = ylim, CI = TRUE)
plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number", ylim = ylim, CI = TRUE)
plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
```

### The results for the Epoch type

```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
op <- par(bty = "n", mfrow = c(3,2))
## Principal colors
colors <- palette()[1:5][-1] #TG: that can be made fancier
colors[5] <- "black"
## Color combination (for the CI)
colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
## y limits
ylim <- range(unlist(lapply(out2$results, function(X) return(c(min(X[,5], na.rm = TRUE),
  max(X[,5], na.rm = TRUE))))))

## Zooming
ylim_zoom <- c(20, 35)

## Plotting the different methods (with/without CIS)
plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration", ylim = ylim, CI = TRUE)
plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number", ylim = ylim, CI = TRUE)
plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number(no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

```

## Plotting the results histogram style

Another way for the plots would be kind of our histogram representation (you can certainly use `ggplot` here, below is just an example):

### The results per method

```{r, fig.width = 12, fig.height = 18}
## Set the list of available models
models <- list("equalbins", "acctran", "deltran", "random", "proximity")

## Get the stratigraphic, duration and number data for the ages
strat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 1)
durat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 2)
numbe_data_ages <- lapply(models, make.class.histogram, out1$results, method = 3)

## Get the stratigraphic, duration and number data for the epochs
strat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 1)
durat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 2)
numbe_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 3)

## The lower age and the lower disparity limit are arbitrarily fixed to 0 and 15 respectively.
lower_age <- 0; lower_lim <- 15
#TG: might want to change that for other datasets!

## limits functions
get.break <- function(x) return(x$breaks)
get.count <- function(x) return(x$counts)
## Ages limits
xlim_ages <- c(max(unlist(c(sapply(strat_data_ages, get.break), sapply(durat_data_ages, get.break),
  sapply(numbe_data_ages, get.break)))), lower_age)
ylim_ages <- c(lower_lim, max(unlist(c(sapply(strat_data_ages, get.count),
  sapply(durat_data_ages, get.count), sapply(numbe_data_ages, get.count)))))
## Epoch limits
xlim_epoc <- c(max(unlist(c(sapply(strat_data_epoc, get.break), sapply(durat_data_epoc, get.break),
  sapply(numbe_data_epoc, get.break)))), lower_age)
ylim_epoc <- c(lower_lim, max(unlist(c(sapply(strat_data_epoc, get.count),
  sapply(durat_data_epoc, get.count), sapply(numbe_data_epoc, get.count)))))

## Colors
colors <- palette()[1:5]
colors[1] <- "grey"

## Graphics
op <- par(bty = "n", mfrow = c(3,2))

## Plot the stratigraphy - ages
plot(strat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - ages")
## Plot the slicing
for(slice in 2:5) {
  lines(strat_data_ages[[slice]]$breaks, strat_data_ages[[slice]]$counts, col = colors[slice])
}

## Adding the legend
legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "random", "proximity"),
  col = colors[1:5], lty = 1, bg = "white")

## Plot the stratigraphy - epoch
plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - Epoch")
## Plot the slicing
for(slice in 2:5) {
  lines(strat_data_epoc[[slice]]$breaks, strat_data_epoc[[slice]]$counts, col = colors[slice])
}

## Plot the duration - ages
plot(durat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Duration - ages")
## Plot the slicing
for(slice in 2:5) {
  lines(durat_data_ages[[slice]]$breaks, durat_data_ages[[slice]]$counts, col = colors[slice])
}

## Plot the duration - epoch
plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Duration - Epoch")
## Plot the slicing
for(slice in 2:5) {
  lines(durat_data_epoc[[slice]]$breaks, durat_data_epoc[[slice]]$counts, col = colors[slice])
}

## Plot the number - ages
plot(numbe_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Number - ages")
## Plot the slicing
for(slice in 2:5) {
  lines(numbe_data_ages[[slice]]$breaks, numbe_data_ages[[slice]]$counts, col = colors[slice])
}

## Plot the number - epoch
plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Number - Epoch")
## Plot the slicing
for(slice in 2:5) {
  lines(numbe_data_epoc[[slice]]$breaks, numbe_data_epoc[[slice]]$counts, col = colors[slice])
}
```

Some of the plots are a bit off set and everything but it gives us an idea

### The results per model

Something like that but with transparency:

```{r, fig.width = 6, fig.height = 6}
plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Epoch")
plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[2], main = "", add = TRUE)
plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[3], main = "", add = TRUE)
```

## Comparing the disparity through time curves

```{r}
curve1 <- out2$object[[1]][[1]]
curve2 <- out2$object[[1]][[2]]
curve3 <- out2$object[[1]][[3]]
curve4 <- out2$object[[1]][[4]]
curve5 <- out2$object[[1]][[5]]

```

>>>>

## Load functions using source
source("../functions/simulations_fun.R")
```

@@

To run `dispRity` analyses we need the **morphospace**, the **phylogeny** (with a root age and node labels) and the **first and last occurrence dates** (FADLAD). The FADLAD is in the `data/` folder, the tree with node labels and cleaned of taxa not in the morphospace is in the `data/processed` folder. To construct the morphospace we need to use the `distance_matrix` files we created in `01-extract-data-for analyses.Rmd` and stored in `data/processed`.

@@

@@

## Loading the tree
## See "01-extract-data-for analyses.Rmd" for how we cleaned these
tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
## Add root age (max tree age date)
tree$root.time <- max(tree.age(tree)[, 1])
```

Ordinating the matrix
----------------

Ordination is a classic NMDS (multidimensional scaling) including only the non-zero variance axes.

```{r}
## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

## Ordinating the matrix
morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
```

@@

These analyses use a series of functions in `functions\`. These essentially wrap the existing `dispRity` functions so that running lots of models and extracting lots of data is not such a big issue. The function `run.all.disparity.wrapper` will run all the scenarios above.

```{r, eval = TRUE, warning = FALSE}
## Run the function for Beck data
slug <- "Beck2014"

## Ages (rarefied to 3 taxa only)
out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
                                bootstraps = 100, metric = c(sum, variances),
                                metric.name = "sum_var", rarefaction = FALSE)

## Save results to output folder
save(out1, file = paste0("../outputs/age.", "Beck2014" ,".Rda"))

## Epochs (rarefied to 3 taxa only)
out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
                        rarefaction = FALSE)

## Save results to output folder
save(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
```
